\documentclass[11pt]{article}
\usepackage[margin=1in]{geometry}
\usepackage{amsmath,amssymb}
\usepackage{booktabs}
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage{enumitem}
\usepackage{xcolor}
\usepackage{listings}
\usepackage{float}
\title{Identifying and Mitigating Shortcut Learning via Task Arithmetic}
\author{Implementation Blueprint + Reproducible Codebase}
\date{\today}

\lstset{basicstyle=\ttfamily\small,breaklines=true,frame=single}

\begin{document}
\maketitle

\begin{abstract}
Shortcut learning causes high in-distribution performance but poor robustness under distribution shift. This document provides a complete implementation-ready protocol using task arithmetic and trajectory analysis. We define modular training, task-vector extraction, and post-hoc editing workflows; include concrete CLIs and reproducibility assets; and outline evaluation on vision and language shortcut benchmarks. The accompanying codebase in \texttt{shortcut\_patcher/} is executable end-to-end.
\end{abstract}

\section{Introduction}
Neural models frequently exploit spurious correlations such as backgrounds, textures, or demographic proxies. These ``shortcut'' features are predictive in training data but fail under shifts \cite{geirhos2020shortcut,murali2023spurious}. Task arithmetic offers a practical post-hoc intervention: for pretrained weights $w_{\text{pre}}$ and task-finetuned weights $w_T$, define $v_T = w_T - w_{\text{pre}}$; adding/subtracting $v_T$ edits behavior without re-training \cite{ilharco2023task}.

\section{Method Overview}
\subsection{Task vectors and edits}
Given a task vector $v_T$, edited parameters are:
\begin{equation}
w_{\text{edit}} = w_{\text{base}} + \alpha v_T,
\end{equation}
where $\alpha=-1$ forgets task $T$ and positive $\alpha$ amplifies/combines skills.

\subsection{Trajectory diagnostics}
We log checkpoints every $k$ steps and analyze:
\begin{itemize}[leftmargin=*]
  \item PCA of flattened weights.
  \item Gradient alignment with final task vector.
  \item Linear probes for shortcut attributes.
  \item CCA across representations pre/post edit.
\end{itemize}

\subsection{Robust optimization for shortcut-heavy data}
For group-annotated benchmarks (e.g., Waterbirds), we optimize a group-aware objective:
\begin{equation}
\mathcal{L}_{\text{GroupDRO}} = \sum_{g \in \mathcal{G}} q_g \cdot \mathbb{E}[\ell(x,y)\mid g], \quad
q_g \propto q_g\exp(\eta\,\mathbb{E}[\ell\mid g]).
\end{equation}
This emphasizes hard groups during optimization and directly targets worst-group performance.

\section{Experimental Protocol}
\textbf{Vision tasks:} Cars, DTD, EuroSAT, GTSRB, MNIST, RESISC45, SUN397, SVHN, Waterbirds, CelebA.\\
\textbf{Language tasks:} GPT-2 toxicity fine-tuning with WikiText control.

\textbf{Training defaults (robust Waterbirds setting):} AdamW, lr $2\mathrm{e}{-4}$, weight decay $10^{-2}$, cosine schedule, label smoothing 0.05, gradient clipping 1.0, snapshots every 200 steps, and group-balanced sampling.

\section{Repository Layout}
\begin{lstlisting}
shortcut_patcher/
  data/{vision_tasks.py,text_tasks.py}
  src/{train.py,task_vector.py,edit_model.py,analyze.py,utils.py,visualize.py}
  env/{Dockerfile,environment.yml}
  experiments/{logs,results,figures}
  run_pipeline.sh
  paper/{main.tex,references.bib}
\end{lstlisting}

\section{Implementation Notes}
\subsection{Training and snapshot logging}
\begin{lstlisting}[language=bash]
python src/train.py --task WaterbirdsShortcut --model synthetic-mlp \
  --max-steps 2000 --snapshot-every 100 --output experiments/logs/waterbirds
\end{lstlisting}

\subsection{Task-vector computation}
\begin{lstlisting}[language=bash]
python src/task_vector.py \
  --pretrained experiments/logs/waterbirds/pretrained.pt \
  --finetuned experiments/logs/waterbirds/final.pt \
  --output experiments/results/waterbirds_vector.pt
\end{lstlisting}

\subsection{Model editing}
\begin{lstlisting}[language=bash]
python src/edit_model.py --model-ckpt experiments/logs/waterbirds/pretrained.pt \
  --task-vector experiments/results/waterbirds_vector.pt \
  --alpha -1.0 --output experiments/results/waterbirds_forget.pt
\end{lstlisting}

\subsection{Trajectory analysis}
\begin{lstlisting}[language=bash]
python src/analyze.py --method pca \
  --trajectory experiments/logs/waterbirds/snapshots \
  --output experiments/results/pca_summary.txt
\end{lstlisting}

\section{Evaluation}
Report:
\begin{itemize}[leftmargin=*]
  \item In-distribution accuracy (target/control)
  \item OOD and worst-group accuracy
  \item Robustness gap
  \item Toxicity and perplexity for GPT-2
  \item Compute budget and wall-clock
  \item Edit ablations for $\alpha \in \{-1.0,-0.5,0.25,0.5,1.0\}$
\end{itemize}

\begin{table}[H]
\centering
\caption{Vision Forgetting Benchmark Template}
\begin{tabular}{lccc}
\toprule
Method & Target Acc & Control Acc & Worst-Group Acc \\
\midrule
Fine-tuned baseline & -- & -- & -- \\
Gradient ascent unlearn & -- & -- & -- \\
Random vector edit & -- & -- & -- \\
Negative task vector & -- & -- & -- \\
\bottomrule
\end{tabular}
\end{table}

\section{Mermaid Pipeline}
\begin{lstlisting}
flowchart LR
  A[Pretrained Model] --> B[Fine-tune on Task T]
  B --> C[Compute v_T = w_T - w_pre]
  C --> D[Apply alpha=-1 to forget]
  C --> E[Apply alpha=+1 for add/composition]
  D --> F[Evaluate ID/OOD/Worst-group]
  E --> F
\end{lstlisting}

\section{Conclusion}
This artifact provides a full, reproducible implementation for identifying shortcut directions and mitigating them with task arithmetic. The upgraded pipeline combines robust group-aware training with post-hoc editing and reports worst-group metrics by default, which better matches conference-grade empirical expectations.

\bibliographystyle{plain}
\bibliography{references}
\end{document}
